---
title: "2023 554 R Notes on Spatial Data Analysis"
author: |
  | Jon Wakefield
  | Departments of Biostatistics and Statistics
  | University of Washington
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  ioslides_presentation: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(RColorBrewer)
library(sf)
library(classInt)
library(tidyverse)
opts_chunk$set(message=FALSE, collapse=TRUE,fig.align='center', tidy.opts=list(blank=TRUE, width.cutoff=70,strip.white=TRUE), warning=FALSE, tidy=TRUE, cache=TRUE)
```

## R for Spatial Analysis

``R`` has extensive spatial capabilities, the 
Spatial task view is [here](http://cran.r-project.org/web/views/Spatial.html) 

Some of the notes that follow are build on Roger Bivand's notes taken from the latter site, and these are based on Bivand et al. (2013), which  is a good reference book of developments until that point.

To get R code alone then load the ``knitr`` library and then type

``purl("2023-554-Spatial-Classes.Rmd")`` 

from the directory with this file in.

### Overview of Spatial Classes

Class definitions are objects that contain the formal definition of a class of R objects, and are usually referred to as an S4 class.

Spatial classes were defined to represent and handle spatial data, so that data can be exchanged between different classes - they are different from regular classes since they need to contain information about spatial locations and their coordinate reference systems

* The ``sp`` library was traditionally the workhorse for representing spatial data.

* Now ``sf`` is being increasingly used, details are [here](https://r-spatial.github.io/sf/)

These notes will focus on ``sf``.

See [Spatial Data Science: With Applications in R](https://r-spatial.org/book/) by Pebesma and Bivand, for a more up to date version.

Another resource is the book [Geocomputation with R](https://geocompr.robinlovelace.net/index.html)

### The ``sf`` Representation

The ``sf`` is the successor package to R packages formerly used in spatial data including ``sp``, ``rgeos`` and the vector parts of ``rgdal``, providing an interface to certain tidyverse packages.

This package reads and writes data through GDAL, and uses GEOS, s2geometry, and PROJ.

The ``sf`` package has ``sf`` objects, a sub-class of a ``data.frame`` or a ``tibble``. These objects contain at least one geometry list-column of class ``sfc``, where each list element contains the geometry as an R object of class ``sfg``. 

All functions are prefixed with `st_` to indicate "spatial type" to make them more easily searchable on the command line.

You can load the package by running `library(sf)`, but note that if you've never used this package before, you should run `install.packages('sf')` in your R console, and then load the package with the `library(sf)` command. 


### Creating a Spatial Object

We can use the `st_read` function (part of the `sf` package) to read in a spatial file. 

```{r}
file <- system.file("gpkg/nc.gpkg", package = "sf")
nc <- st_read(file)
```

Note that the `st_read()` function takes in two arguments, the `dsn`, or data source name, and the layer. Our provided data has only one layer, so that argument is omitted in our example. You can check the available layers by querying:

```{r}
st_layers(file)
```

It's also possible to subset this data and plot it. You can run each line sequentially in R to see what each line of code adds to the plot. Note that this is written in base R, not with ``ggplot()` where each line of code builds upon the plot.

```{r}
# Subset the data
nc5 <- nc[1:5, ]
nc7 <- nc[1:7, ]

# Plot the data
plot(st_geometry(nc7))
plot(st_geometry(nc5), add = TRUE, border = "brown")
cc <- st_coordinates(st_centroid(st_geometry(nc7)))
text(cc, labels = 1:nrow(nc7), col = "blue")
```

### Visualizing Spatial Data

As you can see below, we can create a plot of the sizes in area of each county in North Carolina. The area part comes into play when we index the `nc` object with a 1. If you change this to a 2, you'll see that this now plots the perimeter of each county instead, as perimeters is the second feature.

```{r}
par(mar = c(0,0,1,0))
plot(nc[1], reset = FALSE) # reset = FALSE: we want to add to a plot with a legend
plot(nc[1,1], col = 'grey', add = TRUE)
```

We can also look at the `meuse` dataset from the `sp` package and convert the data into a `sp` object, namely a `SpatialPointsDataFrame` by using the `coordinates()` function. Then, using `st_as_sf()`, we can go from an `sp` object to an `sf` object.

```{r}
data(meuse, package = "sp")
coordinates(meuse) = ~x+y
m.sf = st_as_sf(meuse)
opar = par(mar=rep(0,4))
plot(m.sf)
```

### John Snow Example

For fun, let's look at the poster child of health mapping.

The Snow data consists of the relevant 1854 London streets, the location of 578 deaths from cholera, and the position of 13 water pumps (wells) that can be used to re-create John Snow's map showing deaths from cholera in the area surrounding Broad Street, London in the 1854 outbreak. 

```{r}
library(HistData)
data(Snow.deaths)
data(Snow.pumps)
data(Snow.streets)
# data(Snow.polygons)
```

We first create an `sfc` object containing the coordinates of the streets using  the  `st_linestring()` function; `st_sfc()` combines all the individual `LINESTRING` pieces:

```{r}
# Streets
slist <- split(Snow.streets[,c("x","y")],
            as.factor(Snow.streets[,"street"]))
m <- lapply(slist, as.matrix) |> lapply(st_linestring) |> st_sfc()
plot(m)
```

Display the streets and then add the deaths and pumps (with labels). The red squares are deaths, blue triangles are pumps
 
```{r, fig.height=5}
plot(m, col="gray")
# deaths:
p <- st_as_sf(Snow.deaths, coords = c("x", "y"))
plot(p, add=TRUE, col ='red', pch=15, cex=0.6)
# pumps:
spp <- st_as_sf(Snow.pumps, coords = c("x", "y"))
plot(spp, add=TRUE, col='blue', pch=17, cex=1.0)
text(Snow.pumps[,c("x","y")], labels=Snow.pumps$label, 
       pos=1, cex=0.8)
```

### Reading Shapefiles

ESRI (a company one of whose products is ArcGIS) shapefiles consist of three files, and this is a common form. 

* The first file (*.shp) contains the geography of each shape. 

* The second file (*.shx) is an index file which contains record offsets. 

* The third file (*.dbf) contains feature attributes with one record per feature.

The Washington state Geospatial Data Archive can be accessed
[here](http://wagda.lib.washington.edu/) contains data that can be read into ``R``. 

As an example, consider  ashington county data that was downloaded from wagda.

The data 
consists of the three files: `wacounty.shp`, `wacounty.shx`, `wacounty.dbf`.

The following code reads in these data and then draws a county level
map of 1990 populations, and a map with centroids.

First load the libraries.

```{r}
library(maps) # for background map outlines
```

Note that there are problems with the files, which are sorted by using the ``repair=T`` argument.

The data can be downloaded from here:
[here]("http://faculty.washington.edu/jonno/SISMIDmaterial/wacounty.shp")


```{r}
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/wacounty.shp",destfile = "wacounty.shp")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/wacounty.shx",destfile = "wacounty.shx")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/wacounty.dbf",destfile = "wacounty.dbf")
wacounty = st_read(dsn=".",layer = "wacounty")
class(wacounty)
```

Let's see what these variables look like: we see county names and FIPS codes.

```{r}
names(wacounty)
wacounty$AreaName[1:3] 
wacounty$AreaKey[1:3] 
# head(wacounty)
```

### Drawing a map

We look at some variables.

```{r}
wacounty$INTPTLAT[1:3] # latitude
wacounty$INTPTLNG[1:3] # longitude
wacounty$CNTY[1:3]    
wacounty$TotPop90[1:3]
```

We look at some variables, and then set up the colors to map. We map 1990 Washington population counts by census tracts.

```{r}
# function to make legend; copy from maptools package
leglabs = function(vec, under = "under", over = "over", between = "-") {
    x <- vec
    lx <- length(x)
    if (lx < 3)
        stop("vector too short")
    res <- character(lx - 1)
    res[1] <- paste(under, x[2])
    for (i in 2:(lx - 2)) res[i] <- paste(x[i], between, x[i + 1])
    res[lx - 1] <- paste(over, x[lx - 1])
    res
}
```

```{r}
plotvar <- wacounty$TotPop90 # variable we want to map
summary(plotvar)
nclr <- 8 # next few lines set up the color scheme for plotting
plotclr <- brewer.pal(nclr,"Oranges")
brks <- round(quantile(plotvar,probs=seq(0,1,1/(nclr))),digits=1)
colornum <- findInterval(plotvar,brks,all.inside=T)
colcode <- plotclr[colornum]
plot(st_geometry(wacounty), col=colcode)
legend(-119,46,legend=leglabs(round(brks,digits=1)),fill=plotclr,cex=0.35,bty="n")
```

In ggplot:

```{r}
ggplot(data = wacounty) +
  geom_sf(aes(fill = colcode)) +
  scale_fill_manual(values = rev(plotclr), labels = rev(leglabs(round(brks,digits=1)))) +
  theme_bw() +
  coord_sf() +
  xlab("") +
  ylab("") +
  theme(legend.title = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

As another alternative we can use the ``plot`` function, which uses base plot for spatial data with attributes. We map the 1990 Washington population counts by county.

```{r, fig.height=5.5}
plot(wacounty["TotPop90"], breaks = brks)
```

We can also use `tmap`:

```{r}
library(tmap)
sel = !st_is_empty(wacounty)
tm_shape(st_set_crs(wacounty, 4326)[sel,]) +
    tm_polygons("TotPop90", breaks = brks) +
    tm_layout(legend.outside = TRUE)
```

For illustration, we define our own cutpoints by hand.

```{r,fig.height=5.5, eval=FALSE}
summary(wacounty$TotPop90)
wacounty$cats <- cut(wacounty$TotPop90,breaks=c(0,17000,39000,100000,1520000),
                     labels=c("<17,000","17,000-39,000","39,000-100,000",">100,000"))
tm_shape(st_set_crs(wacounty, 4326)[sel,]) +
    tm_polygons("cats") +
    tm_layout(legend.outside = TRUE)
```

We now highlight a county

```{r, fig.height=5.5}
# identify counties of interest
xx=which(wacounty$CNTY==33)
xx2=which(wacounty$CNTY==63)

# plot the whole state
plot(st_geometry(wacounty), border="#00000075")

# highlight counties of interest
plot(st_geometry(wacounty[xx, ]), col="#ff000075", add=T)
plot(st_geometry(wacounty[xx2, ]), col=NA, border="#0000ff75", add=T, lwd=2.5)

# Add some labels
text(st_coordinates(st_centroid(st_geometry(wacounty[xx, ]))),
     "King", cex=0.75, pos=3, offset=0.25)
text(st_coordinates(st_centroid(st_geometry(wacounty[xx2, ]))),
     "Spokane", cex=0.7, pos=1, offset=0.25)
points(st_coordinates(st_centroid(st_geometry(wacounty[c(xx, xx2), ]))),
       pch=16, cex=0.75)
```

Now let's repeat for census tracts.

```{r}
## Reading Shapefiles
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/watract.shp",destfile = "watract.shp")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/watract.shx",destfile = "watract.shx")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/watract.dbf",destfile = "watract.dbf")
watract=st_read(dsn=".",layer = "watract")
```

Drawing a census tract map: we repeat but now map populations at the census tract level.

```{r, fig.height=5.5}
watract <- st_read("watract.shp") |> st_make_valid() |> st_set_crs(4326)
names(watract)
plotvar <- watract$TotPop90 # variable we want to map
brks <- round(quantile(plotvar,probs=seq(0,1,1/(nclr))),digits=1)
colornum <- findInterval(plotvar,brks,all.inside=T)
colcode <- plotclr[colornum]
plot(st_geometry(watract), col=colcode)
legend(-119,46,legend=leglabs(round(brks,digits=1)),
         fill=plotclr,cex=0.4,bty="n")
```

We zoom in on King County

```{r, fig.height=5.5}
xx=which(watract$CNTY==33)
plot(st_geometry(watract[xx,]), col=colcode[xx])
legend("bottomleft",legend=leglabs(round(brks,digits=1)),
         fill=plotclr,cex=0.4,bty="n")
```

A county map of North Carolina with text

```{r,fig.height=5.5}
library(maps)
map("county", "North Carolina", fill = TRUE, plot = FALSE) |> 
  st_as_sf() |> 
  st_make_valid() -> nc
centroids = st_centroid(st_geometry(nc)) |> st_coordinates()
plot(st_geometry(nc))
text(centroids[,1], centroids[,2], 1:100, offset=0, cex=0.4)
```

In ggplot:

```{r,fig.height=5.5}
ggplot() +
  geom_sf(data = nc, col = "black", fill = NA) +
  geom_text(aes(x = centroids[ , 1], y = centroids[ , 2], 
                label = 1:100), size = 2) +
  theme_bw() +
  coord_sf() +
  xlab("") +
  ylab("") +
  theme(legend.title = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```


