---
title: "2023 554 R Notes on Prevalence Mapping"
author: |
  | Jon Wakefield
  | Departments of Biostatistics and Statistics
  | University of Washington
date: "`r Sys.Date()`"
output: pdf_document
---
  
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(collapse=TRUE, fig.align='center', tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=70,strip.white=TRUE), warning=FALSE,message=FALSE,cache=F)
```

# Overview and Preliminaries

## Overview

In these notes we analyze Demographic Health Survey (DHS) data on HIV prevalence in women age 15-29 in Malawi.

The raw data can be requested from the DHS. Here we use data summaries. 

We will use the ``SUMMER`` package for the computation and plotting.

We will obtain:
  
  * Direct estimates (area-level)

* Smoothed direct estimates (area-level)

* Betabinomial cluster-level estimates (unit-level)


## Libraries 

```{r,fig.width=4.5, fig.height=3.3}
# Install necessary packages 
library(tidyverse)
# install.packages("INLA",repos=c(getOption("repos"),
# INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)
library(INLA)
# devtools::install_github("richardli/SUMMER", build_vignettes = F, force = T)
library(SUMMER)
library(spdep)
library(mapproj)
library(ggpubr)
library(sf)
```


## Read in the data

```{r}
# read in direct estimates and shape files
# must have a folder named R-examples in your current working directory
if (!("R-examples" %in% list.files())) {
  dir.create("R-examples")
}

download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/direct.csv",
              destfile = "R-examples/direct.csv")
direct_df <- read.csv("R-examples/direct.csv")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/gadm36_MWI_1.shx",
              destfile = "R-examples/gadm36_MWI_1.shx")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/gadm36_MWI_1.shp",
              destfile = "R-examples/gadm36_MWI_1.shp")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/gadm36_MWI_1.dbf",
              destfile = "R-examples/gadm36_MWI_1.dbf")
geo <- st_read("R-examples", layer = "gadm36_MWI_1")
```


## Remove a small island

```{r}
# filter out Likoma 
geo <- geo[geo$NAME_1 != "Likoma",] 

# Make lowercase to match DHS data
geo$NAME_1 <- geo$NAME_1 %>% str_to_lower() %>% factor()
geo$id <- rownames(geo)
rownames(geo) <- geo$NAME_1

# create neighbor matrix
nb.r <- spdep::poly2nb(geo, queen = F)
mat <- spdep::nb2mat(nb.r, style = "B", zero.policy = TRUE)
colnames(mat) <- rownames(mat)
```


# Direct Estimates

## Map of direct estimates 


```{r,fig.height=2.3,fig.width=2.3}
mapPlot(direct_df, variables = "p", geo = geo, by.data = "region", 
        by.geo = "NAME_1", legend.label = "Prevalence",
        removetab = TRUE)
```

## Map of uncertainty of direct estimates 

```{r,fig.height=2.3,fig.width=2.3}
mapPlot(direct_df, variables = "p_sd", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "SD",
        removetab = TRUE)
```


## Map of covariate, which is percent of households in that region that have a female head of household


```{r,fig.height=2.3,fig.width=2.3}
mapPlot(direct_df, variables = "hoh", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "Female Head \n of Household",
        removetab = TRUE)
```


# Smoothed Direct

## Smoothed direct estimates using SUMMER

Obtain smoothed direct estimates using SUMMER - BYM2 model for space.

```{r}
# add variance columns to dataframe
dat <- direct_df %>% 
  mutate(logit_p_var = logit_p_sd^2,
         p_var = p_sd^2)

# get smoothed direct estimates from INLA
smoothed <- smoothSurvey(data = NULL, direct.est = dat, 
                         X = dat %>% select(region, hoh), # include covariate
                         Amat = oldmat, regionVar = "region",
                         responseVar = "p", direct.est.var = "p_var",
                         responseType = "binary")
```


## Summaries from the smoothed direct model

```{r}
head(smoothed$HT,3)
smoothed$fit$summary.fixed[,1:5]
smoothed$fit$summary.hyperpar[,1:5]
```

## Map of median of smoothed direct estimates 

```{r, fig.height=2.3,fig.width=2.3}
mapPlot(smoothed$smooth, variables = "median", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "Prevalence",
        removetab = TRUE)
```


## Map of uncertainty of smoothed direct estimates 

```{r, fig.height=2.3,fig.width=2.3}
smoothed$smooth$sd <- smoothed$smooth$var %>% sqrt()
mapPlot(smoothed$smooth, variables = "sd", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "SD",
        removetab = TRUE)
```


# Betabinomial Model

## Betabinomial model: load the data


```{r}
# read in cluster-level data and urban/rural proportions at area level
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/urban_prop_sample_df.csv",
              destfile = "R-examples/urban_prop_sample_df.csv")
download.file("http://faculty.washington.edu/jonno/SISMIDmaterial/binom.csv",
              destfile = "R-examples/binom.csv")
binom_df <- read.csv("R-examples/binom.csv")
urb_prop_df <- read.csv("R-examples/urban_prop_sample_df.csv", stringsAsFactors = FALSE)
binom_df$strata_within_region <- ifelse(binom_df$urban == 1, "urban", "rural")
weights <- urb_prop_df %>% 
  mutate(urban = prop,
         rural = 1 - prop) %>% 
  select(region, urban, rural)
```


## Sort out the covariate and fit the model


```{r}
# create covariate data frame for input to smoothSurvey()
# covariate is at the area level
X <- binom_df %>% distinct(region, .keep_all = TRUE) %>% dplyr::select(region, head_house)

bb <- smoothSurvey(data = binom_df, Amat = mat, X = X,
                   is.unit.level = TRUE, is.agg = TRUE,
                   responseType = "binary", responseVar = "y", regionVar = "region", 
                   weightVar = NULL, strataVar.within = "strata_within_region", 
                   clusterVar = "clustid", totalVar = "n",
                   weight.strata = weights, nsim = 1000)

bb$smooth.overall$sd <- bb$smooth.overall$var %>% sqrt()
```


## Summaries from the betabinomial model

```{r}
bb$fit$summary.fixed[,1:5]
bb$fit$summary.hyperpar[,1:5]

```

## Betabinomial estimates map

```{r, fig.height=2.3,fig.width=2.3}
mapPlot(bb$smooth.overall, variables = "median", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "Prevalence",
        removetab = TRUE)

```


## Betabinomial uncertainty map

```{r, fig.height=2.3,fig.width=2.3}
mapPlot(bb$smooth.overall, variables = "sd", geo = geo, by.data = "region", 
        by.geo = "NAME_1", direction = -1, legend.label = "SD",
        removetab = TRUE)
```


## Combine point estimates

```{r}
# define prevalence plots
direct_prev <- mapPlot(direct_df, variables = "p", geo = geo, by.data = "region", 
                       by.geo = "NAME_1", direction = -1, legend.label = "Prevalence",
                       removetab = TRUE)
smooth_prev <- mapPlot(smoothed$smooth, variables = "median", geo = geo, by.data = "region", 
                       by.geo = "NAME_1", direction = -1, legend.label = "Prevalence",
                       removetab = TRUE)
bb_prev <- mapPlot(bb$smooth.overall, variables = "median", geo = geo, by.data = "region", 
                   by.geo = "NAME_1", direction = -1, legend.label = "Prevalence",
                   removetab = TRUE)
```


## Combine uncertainty estimates

```{r}
# define sd plots
direct_sd <- mapPlot(direct_df, variables = "p_sd", geo = geo, by.data = "region", 
                     by.geo = "NAME_1", direction = -1, legend.label = "SD",
                     removetab = TRUE)
smooth_sd <- mapPlot(smoothed$smooth, variables = "sd", geo = geo, by.data = "region", 
                     by.geo = "NAME_1", direction = -1, legend.label = "SD",
                     removetab = TRUE)
bb_sd <- mapPlot(bb$smooth.overall, variables = "sd", geo = geo, by.data = "region", 
                 by.geo = "NAME_1", direction = -1, legend.label = "SD",
                 removetab = TRUE)

```


## Comparison of point estimates across methods

```{r, fig.height=2.6,fig.width=4.6}
ggarrange(direct_prev, smooth_prev, bb_prev, nrow = 1, common.legend = TRUE, legend = "right")
```


## Comparison of uncertainty across methods


```{r, fig.height=2.6,fig.width=4.6}
ggarrange(direct_sd, smooth_sd, bb_sd, nrow = 1, common.legend = TRUE, legend = "right")
```

