---
title: "2023 554 SUMMER Package R Notes"
author: |
  | Jon Wakefield
  | Departments of Biostatistics and Statistics
  | University of Washington
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(collapse=TRUE, fig.align='center', tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=60,strip.white=TRUE), warning=FALSE,message=FALSE,cache=FALSE)
```

## Small Area Estimation (SAE)

In these notes, SAE via the ``SUMMER`` package will be illustrated.

Details on ``SUMMER``, including a vignette, can be found [here](https://cran.r-project.org/web/packages/SUMMER/index.html)


We illustrate with the Washington State BRFSS diabetes example and will obtain:

- Naive estimates
- Weighted estimates
- Estimates from a binomial BYM2 model 
- Estimates from Fay-Herriot models

## Load ``SUMMER`` package


We first load the ``SUMMER`` package.

``INLA`` is not in a standard repository, so we check if it is
available and install it if it is not.


```{r}
library(SUMMER)
if (!isTRUE(requireNamespace("INLA", quietly = TRUE))) {
install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
}
```

## Read in Data

``BRFSS`` contains the full BRFSS dataset with 16,283 observations:

- ``diab2`` variable is the binary indicator
of Type II diabetes

- ``strata`` is the strata indicator and 

- ``rwt_llcp`` is the final  weight. 

For the purpose
of this analysis, we first remove records with missing HRA code or diabetes status from this dataset.


```{r}
data(BRFSS)
BRFSS <- subset(BRFSS, !is.na(BRFSS$diab2))
BRFSS <- subset(BRFSS, !is.na(BRFSS$hracode))
```


``KingCounty`` contains the map of the King County HRAs. In order to fit spatial smoothing model, we first need to compute the adjacency matrix for the HRAs, ``mat``, and make sure both the column and row names correspond to the HRA names.

```{r}
library(sf) # Load sf for spatial analysis
library(prioritizr) # Allows us to create an adjacency matrix

data(KingCounty)
KingCounty <- st_as_sf(KingCounty)
nb.r <- st_centroid(st_geometry(KingCounty), of_largest_polygon=TRUE)
mat <- adjacency_matrix(KingCounty)
colnames(mat) <- rownames(mat) <- KingCounty$HRA2010v2_
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])
mat[1:2,1:2]
```

## Create ``survey`` object

We load the survey package and then define the survey object for the BRFSS data.
We have stratified, disproportionate sampling, so note the arguments:

- ``weights``
- ``strata``

We then calculate the direct (weighted) esimates.


```{r,}
library(survey)
design <- svydesign(ids = ~1, weights = ~rwt_llcp, 
                    strata = ~strata, data = BRFSS)
direct <- svyby(~diab2, ~hracode, design, svymean)
head(direct,n=7)
```

## Binomial spatial smoothing model

We ignore the design and fit the model:
\begin{eqnarray*}
y_i | p_i &\sim& \mbox{Binomial}(n_i,p_i)\\
\theta_i &=& \log \left( \frac{p_i}{1-p_i}\right) = \alpha + b_i 
\end{eqnarray*}
with $b_i$ following a BYM2 model, i.e., an iid normal random effect and an 
intrinsic CAR (ICAR) random effect.

The binomial smoothing model is fitted by specifying NULL for the survey characteristics.

## The ``smoothSurvey`` function

Note how the polygon information is input, and the neighbors in the ``Amat`` argument - this is required for the ICAR.

```{r}
smoothed <- smoothSurvey(data = BRFSS, geo = KingCounty, Amat = mat, responseType = "binary",
responseVar = "diab2", strataVar = NULL, weightVar = NULL, regionVar = "hracode",
clusterVar = NULL, CI = 0.95)
```

The usual INLA summaries can be found in ``smoothed$fit``:
```{r}
smoothed$fit$summary.fixed
smoothed$fit$summary.hyper
```

Now examine some of the other components:
```{r}
names(smoothed)
names(smoothed$HT)
names(smoothed$smooth)
head(smoothed$HT,n=4)
```

The smoothed estimates of $p_i$ and $\theta_i$ can be found in the smooth object returned by the function, and the
direct estimates are stored in the HT object (without specifying survey weights, these are the simple binomial
probabilities).


```{r, echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff=50), fig.height=1.5, fig.width = 3, eval = TRUE}
head(smoothed$smooth,n=1)
head(smoothed$HT,n=1)
```


We map the posterior mean estimates.

```{r}
data(KingCounty)
toplot <- smoothed$smooth
mapPlot(data=toplot,geo=KingCounty,
variables=c("mean"),
labels=c("Posterior Mean"),by.data="region",by.geo="HRA2010v2_")
```


We map the Direct Estimates, which are available in the ``smoothSurvey`` fit.


```{r, echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff=50), fig.height=2.2, fig.width = 4, eval = TRUE}
toplot$HTest <- smoothed$HT$HT.est
mapPlot(data=toplot,geo=KingCounty,
variables=c("HTest"),
labels=c("Direct Estimates"),by.data="region",by.geo="HRA2010v2_")
```

Now map the lower and upper endpoints of 95\% CI for direct estimates


```{r, fig.height=2,fig.width=6 }
lo <- smoothed$HT$HT.est-1.96*sqrt(smoothed$HT$HT.var)
hi <- smoothed$HT$HT.est+1.96*sqrt(smoothed$HT$HT.var)
toplot$HTlower <- lo
toplot$HTupper <- hi
mapPlot(data=toplot,geo=KingCounty,
variables=c("HTlower","HTupper"),
labels=c("Direct Lower","Direct Upper"),by.data="region",by.geo="HRA2010v2_")
```

## Fit Fay-Herriot smoothing model, which  acknowledges the design

We now acknowledge the design and fit the model
$$\widehat \theta_i \sim \mbox{N}(\theta_i,\widehat{V}_i)$$
with $\widehat \theta_i = \log[\widehat{p}_i/(1-\widehat{p}_i)]$ where $\widehat p_i$ being the direct estimate and $\widehat V_i$ the variance of this estimate (where the design is acknowledged in the variance calculation) and
$$
\theta_i = \log \left( \frac{p_i}{1-p_i}\right) = \mu + \epsilon_i + \epsilon_i 
$$
with $\epsilon_i \sim_{iid} N(0,\sigma^2)$.

We put ``Amat=NULL`` to obtain an iid model only (i.e., the standard Fay-Herriot model without covariates).

```{r}
FHmodel <- smoothSurvey(data = BRFSS, geo = KingCounty, Amat = NULL, responseType =  "binary",responseVar = "diab2", strataVar = "strata", weightVar = "rwt_llcp",
regionVar = "hracode", clusterVar = "~1", CI = 0.95)
FHmodel$fit$summary.fixed[1:5]
FHmodel$fit$summary.hyper[1:5]
sqrt(1/FHmodel$fit$summary.hyper[3:5])
```

Now extend the random effects structure to allow for BYM2 random effects.

```{r}
svysmoothed <- smoothSurvey(data = BRFSS, geo = KingCounty, Amat = mat, responseType =  "binary",responseVar = "diab2", strataVar = "strata", weightVar = "rwt_llcp",
regionVar = "hracode", clusterVar = "~1", CI = 0.95)
svysmoothed$fit$summary.fixed[1:5]
svysmoothed$fit$summary.hyper[1:2,1:5]
sqrt(1/svysmoothed$fit$summary.hyper[1,3:5])
```

```{r}
est <- data.frame(naive = smoothed$HT$HT.est,
weighted = svysmoothed$HT$HT.est,
smooth = smoothed$smooth$mean,
weightedsmooth = svysmoothed$smooth$mean)
var <- data.frame(naive = smoothed$HT$HT.var,
weighted = svysmoothed$HT$HT.var,
smooth = smoothed$smooth$var,
weightedsmooth = svysmoothed$smooth$var)
```

Setting Up

```{r}
l1 <- range(est)
l2 <- range(var)
library(ggplot2)
g1 <- ggplot(est, aes(x = naive, y = smooth)) + geom_point() +
geom_abline(slope = 1, intercept = 0, color = "red") +
ggtitle("Naive Ests") + xlab("Direct") + ylab("Smoothed")+ xlim(l1) + ylim(l1) 
g2 <- ggplot(var, aes(x = naive, y = smooth)) + geom_point() +
geom_abline(slope = 1, intercept = 0, color = "red") +
ggtitle("Naive Vars") + xlab("Direct") + ylab("Smoothed") + xlim(l2) + ylim(l2)
g3 <- ggplot(est, aes(x = weighted, y = weightedsmooth)) + geom_point() +
geom_abline(slope = 1, intercept = 0, color = "red") +
ggtitle("Survey Wtd Ests") + xlab("Direct") + ylab("Smoothed") + xlim(l1) + ylim(l1)
g4 <- ggplot(var, aes(x = weighted, y = weightedsmooth)) + geom_point() +
geom_abline(slope = 1, intercept = 0, color = "red") +
ggtitle("Survey Wtd Vars") + xlab("Direct") + ylab("Smoothed") + xlim(l2) + ylim(l2)
library(gridExtra)
```

```{r}
grid.arrange(grobs = list(g1, g2, g3, g4), ncol = 2)
```

We can also compare posterior medians of the two Bayes Fay Harriot models (`FHmodel` and `svysmoothed`).

```{r}
p1 <- mapPlot(data=FHmodel$smooth, geo=KingCounty,
        variables = "median",
        labels = "Direct Posterior Median",
        by.data="region",
        by.geo="HRA2010v2_")
p2 <- mapPlot(data=svysmoothed$smooth, geo=KingCounty,
        variables = "median",
        labels = "Smoothed Posterior Median",
        by.data="region",
        by.geo="HRA2010v2_")
grid.arrange(grobs = list(p1, p2), ncol = 2)
```

It's also useful to plot the estimates against each other and the posterior standard errors against each other.

```{r}
# Posterior Estimates
ggplot(
  data.frame(
    smoothed1 = FHmodel$smooth$median,
    smoothed2 = svysmoothed$smooth$median
  ),
  aes(x = smoothed1, y = smoothed2)
) +
  geom_point() +
  labs(y = "Spatial FH Estimate", x = "Non-Spatial FH Estimate") +
  geom_abline(color = "red") +
  coord_equal(xlim = c(0.03, 0.17), ylim = c(0.03, 0.17))

# Posterior SDs
ggplot(
  data.frame(
    smoothed1 = sqrt(FHmodel$HT$HT.var),
    smoothed2 = sqrt(svysmoothed$HT$HT.var)
  ),
  aes(x = smoothed1, y = smoothed2)
) +
  geom_point() +
  labs(y = "Spatial FH Estimate", x = "Non-Spatial FH Estimate") +
  geom_abline(color = "red") +
  coord_equal(xlim = c(0.005, 0.06), ylim = c(0.005, 0.06))
```


## SAE in Space and Time

When data consist of observations from different time periods, we can extend the framework to smooth
estimates over both space and time. The space-time interaction terms are modeled by the type I-IV
interactions -- see Held (2000, Statistics in Medicine).



```{r}
svysmoothed.year <- fitGeneric(data = BRFSS, geo = KingCounty, Amat = mat,
responseType = "binary", responseVar = "diab2", strataVar = "strata", weightVar = "rwt_llcp",
regionVar = "hracode", clusterVar = "~1", timeVar = "year", time.model = "rw1",
type.st = 1)
```

Maps of Posterior Means over Time

```{r}
mapPlot(data = svysmoothed.year$smooth, geo = KingCounty, values = "mean",
variables = "time", by.data = "region", by.geo = "HRA2010v2_", is.long = TRUE)
```

## Final Comments

More materials can be found [here](http://faculty.washington.edu/jonno/index.html)

```SUMMER``` has a Github page with the latest changes, see also:

[paper](https://arxiv.org/pdf/2007.05117.pdf)


